---
title: "COVID-19 UK introductions"
subtitle: "Extract UK transmission lineages (BEAST + DTA)"
author: "Louis du Plessis"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output: 
  github_document:
    toc: true
    toc_depth: 2
layout: page
editor_options: 
  chunk_output_type: inline
params: 
  inputpath  :  "../../../data/phylogenetic/"
  intreepath : "../results/xml/dta/output/"
  outputpath : "../results/combined_beast_dta/"
  cluster_f  : DTA
  nreps      : 2000
  minsize    : 100
  ncores     : 4
  uklimit    : "2020-01-23"
  
---

```{r rsetup, include=FALSE}
  
    library(tictoc)
    library(dplyr)
    library(lubridate)
    library(ape)
    library(tidytree)
    library(parallel)
    library(beastio)
    library(ggplot2)
    source("../scripts/treeutils.R")
    source("../scripts/clusterutils.R")

    inputpath  <- params$inputpath
    outputpath <- params$outputpath
    intreepath <- params$intreepath
    cluster_f  <- params$cluster_f
    replicates <- params$nreps
    minsize    <- params$minsize
    ncores     <- params$ncores
    uklimit    <- decimal_date(as.Date(params$uklimit))  
    


    figpath    <- paste0(outputpath, "figures/")
    cachepath  <- paste0(outputpath, "cache/")
    dir.create(outputpath, recursive = TRUE, showWarnings = FALSE)
    dir.create(figpath, recursive = TRUE, showWarnings = FALSE)

    knitr::opts_chunk$set(tidy=FALSE, cache=FALSE, cache.path = cachepath, cache.lazy = FALSE,
                          dev="png", dpi=150, fig.path=figpath, 
                          message=FALSE, error=FALSE, warning=TRUE, echo=FALSE)

    metadata              <- read.csv(paste0(inputpath, "metadata.csv"))
    metadata$sample_date  <- ymd(metadata$sample_date)
    metadata$decimal_date <- decimal_date(metadata$sample_date)    
    metadata$taxon_label  <- metadata$sequence_name
    #metadata$taxon_label <- gsub("/", "_", as.character(metadata$sequence_name))
    
    treefiles <- list.files(path = intreepath, pattern = "*.trees$")
    
```

```{r functions} 


extractTreeClusters <- function(i, tree, metadata, treefile, treepath, maxheight, minsize, saveTrees=TRUE) {
  
    # Extract clusters
    prefix    <- paste0(strsplit(treefile, "-")[[1]][1], "_",cluster_f,"_")
    treetable <- getClusterTable(tree, "UK", prefix = prefix, maxheight = maxheight, debug = FALSE) 
    write.csv(treetable, paste0(treepath, gsub("\\.trees", "_", treefile), i, ".treetable.csv"), quote=FALSE, row.names=TRUE)
    
    # Should remove singletons?
    clusterSamples      <- treetable[treetable$types == "sample", ]
    clusterMeta         <- droplevels(metadata[match(clusterSamples$taxon, metadata$taxon_label), ])
    clusterMeta$cluster <- clusterSamples$cluster[match(clusterMeta$taxon_label, clusterSamples$taxon)]
    
    # Save summary file
    clusterSizes   <- table(clusterMeta$cluster)
    clusterNames   <- names(clusterSizes)[clusterSizes > 1]
    clusterSingles <- names(clusterSizes)[clusterSizes == 1]
    
    mostrecent <- max(clusterMeta$decimal_date)
    
    tmrcas  <- sapply(clusterNames, function(x) mostrecent - max(treetable$heights[treetable$cluster == x], na.rm = TRUE))
    oldest  <- sapply(clusterNames, function(x) min(clusterMeta$decimal_date[clusterMeta$cluster == x], na.rm = TRUE))
    newest  <- sapply(clusterNames, function(x) max(clusterMeta$decimal_date[clusterMeta$cluster == x], na.rm = TRUE))
    seqs    <- sapply(clusterNames, function(x) sum(clusterMeta$cluster == x, na.rm = TRUE))
    
    treeClusterStats <- data.frame(cluster=clusterNames, seqs=seqs, tmrca=tmrcas, tmrca_calendar=round_date(date_decimal(tmrcas), unit="day"), 
                                         oldest=oldest, mostrecent=newest, treefile=treefile, tree=i)
    
    if (saveTrees) {
        # Extract and save trees for all clusters with >minsize tips
        clusterpath <- paste0(treepath, gsub("\\.trees", "_", treefile), i, "/")
        dir.create(clusterpath, recursive = TRUE, showWarnings = FALSE)
        
        bigClusters <- names(seqs)[which(seqs >= minsize)]
        for (cluster in bigClusters) {
          clusterIds  <- as.character(na.omit(clusterSamples$taxon[clusterSamples$cluster == cluster,drop=TRUE]))
          clusterMRCA <- getMRCA(tree@phylo, clusterIds)
          
          #clustertable <- treetable[treetable$cluster == cluster & !is.na(treetable$cluster), ]
          #clustertable <- clustertable[order(clustertable$heights, decreasing=TRUE), ]
          #clusterMRCA2 <- as.numeric(rownames(clustertable)[1])
          
          clusterTree  <- treeio::tree_subset(tree, clusterMRCA, levels_back=0)
          #write.tree(clusterTree@phylo, paste0(clusterpath, cluster, ".tree"))
          write.beast(clusterTree, paste0(clusterpath, cluster, ".tree"))
          
        }
    }
    
    return(treeClusterStats)
}
  


```



# Summary

This notebook extracts TMRCAs and other summary statistics of the `r cluster_f` clusters across all replicate trees from BEAST.

## Input
- Metadata table in `.csv` format. Should contain a `sequence_name` and `sample_date` column.
- Set of trees for each treefile with DTA reconstructions save in `r intreepath`.

## Steps
- Extract a table of nodes and heights for each lineage in each tree. 
- Extract TMRCAs and sizes for each lineage in each tree.
- Extract subtrees for all lineages with at least `r minsize` tips.
- Extract lineage assignment for all UK sequences in each tree replicate.

## Output
- `.csv` files with nodes and heights for each lineage in each tree.
- `.csv` file with lineage assignments for all UK sequences in each tree replicate. 
- `.Rds` files for each tree (easier to load, since it's already a treedata data structure). 
- Newick trees for all lineages with at least `r minsize` tips. 




# Extract clusters

```{r extractClusters, results="asis", warning=FALSE, message=FALSE, cache=TRUE, eval=TRUE}
   
          
    #tic(cat(paste0("\n\nTotal time for ", replicates, " trees: ", replicates)))
    clusterStats <- c()
    for (treefile in treefiles) {
  
        tic(cat(paste0("\n\n**Loading ", treefile, ":** ")))
        trees <- read.beast.groups(paste0(intreepath, treefile), groupsize = 50, skip=0, ntrees=replicates, debug=FALSE)
        treemeta   <- droplevels(metadata[match(trees[[1]]@phylo$tip.label, metadata$taxon_label), ])
        mostrecent <- max(treemeta$decimal_date)
        toc()
        
        cat(paste0("\n\n - Most recent tip: ", format.Date(round_date(date_decimal(mostrecent), unit = "day")), " (", mostrecent, ")\n"))
        cat(paste0(" - Maximum UK node height: ", mostrecent - uklimit, "\n\n"))
        
        tic(cat("\n - Saving and compressing: "))
        saveRDS(trees, file=paste0(outputpath, treefile, ".Rds"), compress = TRUE)
        toc()
        
        tic(cat("\n - Extract clusters from trees: "))
        treepath <- paste0(outputpath, gsub("\\.trees", "", treefile), "/")
        dir.create(treepath, recursive = TRUE, showWarnings = FALSE)
        
        statList <- mclapply(seq_len(length(trees)), function(i) extractTreeClusters(i, trees[[i]], metadata, treefile, treepath, (mostrecent - uklimit), minsize), mc.cores = ncores)
        
        treeClusterStats <- statList[[1]]
        for (i in 2:length(statList)) {
            treeClusterStats <- rbind(treeClusterStats, statList[[i]])
        }
        
        write.csv(treeClusterStats, paste0(outputpath, gsub("\\.trees", "_", treefile), cluster_f,".csv"), row.names=FALSE, quote=FALSE)
        clusterStats <- rbind(clusterStats, treeClusterStats)
        
        toc()
            
    }
    write.csv(clusterStats, paste0(outputpath, "clusters_",cluster_f,".csv"), row.names=FALSE, quote=FALSE)
    #toc()

    
    
```

# Extract cluster samples

```{r extractClusterSamples, results="asis", eval=TRUE}
  
    treedirs <- gsub("\\.trees", "", treefiles)  
  
    allSamples <- c()
    for (dir in treedirs) {   
        tic(cat(paste0("\n\nProcessing ", dir, ": ")))
      
        treesamples <- NULL
        treetables  <- list.files(path = paste0(outputpath, dir, "/"), pattern = "*.treetable.csv$")
        reps        <- c()
        for (i in 1:length(treetables)) {
            rep  <- treetables[i] %>% sub(".treetable.csv", "", .) %>% sub("_", "", .) %>% sub(dir, "", .) %>% as.numeric
            reps <- c(reps, rep)
            
            treetable        <- read.csv(paste0(outputpath, dir, "/", treetables[i]))
            ukseqs           <- droplevels(treetable[treetable$location == "UK" & treetable$types == "sample", c("taxon", "cluster")])
            colnames(ukseqs) <- c("taxon", paste0("tree", rep))
            
            if (is.null(treesamples)) {
                treesamples <- ukseqs[order(ukseqs$taxon), ]
            } else {
                # Reorder
                ukseqs <- ukseqs[match(treesamples$taxon, ukseqs$taxon), ]
                            
                # Check and bind
                if (all(ukseqs$taxon == treesamples$taxon)) {
                    treesamples[[paste0("tree",rep)]] <- ukseqs[[paste0("tree",rep)]]
                } else {
                   stop(sprintf("Tree %d contains different UK taxa", rep))
                }
                
            }
        }
        allSamples <- rbind(allSamples, treesamples[, c(1, order(reps)+1)])
        toc()
    }
    write.csv(allSamples, paste0(outputpath, "clusterSamples_",cluster_f,".csv"), quote=FALSE, row.names=FALSE)


```


# Session info

```{r sessionInfo, results='markup'}
    sessionInfo()
```
